\thetitle

### 2013-09-29 tfpie/tfpie.tex
Some background on HtDP.

More significantly, Northeastern distinctively emphasizes experiential
education, with almost all Computer Science majors participating in a
6 month ``co-op'' internship after their third semester.  These co-ops
take place at a wide variety of companies, and while most students do
some software development, there is almost no uniformity beyond that.
In particular, there is wide variation in the languages students use
while on co-op.

This combination sets the constraints under which we designed our
approach.  Our students begin the course with a firm grasp of
data-driven program design, as well as experience with the tools used
in the first course:

\begin{itemize}
  \item the student languages introduced in \emph{How to Design
    Programs}~\cite{dvanhorn:Felleisen2004Structure,
    dvanhorn:Felleisen2001How} as well as the idea of ``language levels'';

  \item the DrRacket (formerly DrScheme) 
    programming environment~\cite{dvanhorn:Findler2002DrScheme};

  \item the \texttt{check-expect} testing framework~\cite{local:check-expect};

  \item the ``World'' approach to developing interactive animations
    and games using functional programming and functional
    graphics~\cite{dvanhorn:Felleisen2009Functional,
      local:barland-sfp10}.
\end{itemize}

\noindent
We deliberately reuse all of these elements in our approach.

After our course, our students should both (a) be prepared for
subsequent courses in the curriculum, which expect familiarity with
Java and standard Java libraries, (b) be prepared for co-ops in which
they will use professional-grade languages and tools which will
almost certainly be object-oriented.  More significantly, we aim to
teach the key insights behind the object-oriented approach to program
design.

>>>>

More significantly, Northeastern distinctively emphasizes experiential
education, with almost all Computer Science majors participating in a
6 month ``co-op'' internship after their third semester.  These co-ops
take place at a wide variety of companies, and while most students do
some software development, there is almost no uniformity beyond that.
In particular, there is wide variation in the languages students use
while on co-op. This combination sets the constraints under which we
designed our approach.

\subsection{A first course on \emph{How to Design Programs}}

In the first semester, students are introduced to the ``design
recipe'', a step-by-step process for going from English descriptions
of problems to working programs.  The design recipe involves six steps:
\begin{enumerate}
\item Analyze the information involved in the problem and express
  how to represent it as data.
\item Write down a function signature, a summary of the purporse
  of the function, and a function stub.
\item Illustrate the signature and the purpose statement with some
  functional examples.
\item Take an inventory of the input data that can be used to compute
  an answer.
\item Write the code for the function.
\item Verify the behavior of the program against the functional
  examples given earlier.
\end{enumerate}

Students explore program design using this process in the context of a
series of successively richer pedagogical programming language
levels~\cite{dvanhorn:Felleisen2004Structure,
  dvanhorn:Felleisen2001How} that are included in the DrRacket
(formerly DrScheme) programming
environment~\cite{dvanhorn:Findler2002DrScheme}.
%
The language and environment include several tools in support of the
design recipe.  For example, functional examples can be written as
executable tests by writting \texttt{check-expect}
expression~\cite{local:check-expect}; an algebraic stepper and REPL
are provided to interact with programs at each step of the design
process.

Finally, the first semester course makes extensive use of a library
for developing interactive animations and games using functional
programming and functional
graphics~\cite{dvanhorn:Felleisen2009Functional, local:barland-sfp10}.

\subsection{The goal of the second course}

After the second course, students should both (a) be prepared for
subsequent courses in the curriculum, which expect familiarity with
Java and standard Java libraries, (b) be prepared for co-ops in which
they will use professional-grade languages and tools which will almost
certainly be object-oriented.  More significantly, we aim to teach the
key insights behind the object-oriented approach to program design.
